---
title: "Stat545"
author: "Mark Kinyanjui"
date: "1/20/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning new techniques Stat 545

https://stat545.com/basic-data-care.html


Laoding all the Packages needed

```{r}
install.packages("gapminder")
install.packages("tidyverse")
library(gapminder)
library(tidyverse)


```


## Exploring the structures and datasets 

Different ways to explore a data frame, tibble 

```{r}
str(gapminder)

class(gapminder)

gapminder

head(gapminder)

tail(gapminder)

as_tibble(iris)

names(gapminder)

summary(gapminder)


# Simple plot jus to visualize dataset 

plot(lifeExp ~ year, gapminder)


plot(lifeExp ~ gdpPercap, gapminder)

plot(lifeExp ~ log(gdpPercap), gapminder)

head(gapminder$lifeExp)

summary(gapminder$lifeExp)

hist(gapminder$lifeExp)

class(gapminder$continent)

summary(gapminder$continent)

levels(gapminder$continent)


nlevels(gapminder$continent)


```

# Slowly diving deeper into the learning 

```{r}
table(gapminder$continent)

barplot(table(gapminder$continent))


```

# Using Dplyr  
It is part of the tidyverse package,

Cmd+Shift+M (Mac).- shortcut for %>%

```{r}
filter(gapminder, lifeExp < 29)

filter(gapminder, country == "Rwanda" , year >1979)

filter(gapminder, country %in% c("Rwanda", "Afghanistan"))


gapminder %>% 
  select(year, lifeExp) %>% 
  head(4)

gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(year, lifeExp)


### Creating a copy of gapminder so no changes are made to the original dataset

(my_gap <- gapminder)

my_gap %>% 
  mutate(gdp = pop * gdpPercap)

```

### Replicating data in a row to fit number of levels 
Maybe it would be more meaningful to consumers of my tables and figures to stick with GDP per capita. But what if I reported GDP per capita, relative to some benchmark country. Since Canada is my adopted home, I’ll go with that.

I need to create a new variable that is gdpPercap divided by Canadian gdpPercap, taking care that I always divide two numbers that pertain to the same year.

How I achieve this:

1.Filter down to the rows for Canada.
2.Create a new temporary variable in my_gap:
3.Extract the gdpPercap variable from the Canadian data.
4.Replicate it once per country in the dataset, so it has the right length.
5.Divide raw gdpPercap by this Canadian figure.
6.Discard the temporary variable of replicated Canadian gdpPercap



summarize_at() applies the same summary function(s) to multiple variables

```{r}
my_gap <- gapminder
ctib <- my_gap %>% 
  filter(country == "Canada") 

my_gap <- my_gap %>% 
  mutate (tmp = rep(ctib$gdpPercap, nlevels(country)),
         gdpPercapRel = gdpPercap / tmp,
         tmp = NULL)
head(my_gap)

summary (my_gap$gdpPercapRel)


### Select 

my_gap %>% 
  filter(country == "Burundi", year > 1996) %>% 
  select( yr = year, lifeExp, gdpPercap) %>% 
  select(gdpPercap, everything())


#How many observations do we have per continent?
my_gap %>% 
  group_by(continent) %>% 
  summarize(n = n())

# The tally() function is a convenience function that knows to count rows. It honors groups

my_gap %>% 
  group_by(continent) %>% 
  tally()

# The count() function is an even more convenient function that does both grouping and counting

my_gap %>% 
  count(continent)


#What if we wanted to add the number of unique countries for each continent?

my_gap %>% 
  group_by(continent) %>% 
  summarize ( n = n(),
              n_countries = n_distinct(country)) 

my_gap %>% 
  group_by(continent) %>% 
  summarize(avg_lifeExp = mean(lifeExp))


# Let’s compute average and median life expectancy and GDP per capita by continent by year…but only for 1952 and 2007.
my_gap %>% 
  filter(year %in% c(1952,2007)) %>% 
  group_by(continent, year) %>% 
  summarize_at(vars(lifeExp,gdpPercap), list(~mean(.),~median(.)))


#Let’s focus just on Asia. What are the minimum and maximum life expectancies seen by year?

my_gap %>% 
  filter(continent == "Asia") %>% 
  group_by(year) %>% 
  summarize ( min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp))

```
# Computing with group-wise summaries
Let’s make a new variable that is the years of life expectancy gained (lost) relative to 1952, for each individual country. We group by country and use mutate() to make a new variable. The first() function extracts the first value from a vector. Notice that first() is operating on the vector of life expectancies within each country group.

```{r}
my_gap %>% 
  group_by(country) %>% 
  select(country, year, lifeExp) %>% 
  mutate(lifeExp_gain = lifeExp - first(lifeExp)) %>% 
  filter(year < 1963)


my_gap %>%
  filter(continent == "Asia") %>% 
  select(year, country, lifeExp) %>% 
  group_by(year) %>% 
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2 ) %>% 
  arrange(year) %>% 
  print(n = Inf)


##  which country experienced the sharpest 5-year drop in life expectancy

my_gap %>% 
  select(country, year, continent, lifeExp) %>%
  group_by(continent, country) %>%
  mutate( le_delta = lifeExp - lag(lifeExp)) %>%
  summarize( worst_le_delta = min(le_delta, na.rm = TRUE)) %>% 
  top_n(-1 , wt = worst_le_delta) %>% 
  arrange(worst_le_delta)



```
### Chapter  8 - TidyR
```{r}


```


### Chapter 9 - Writing and reading data 

````{r}
library(tidyverse)
library(fs)
(gap_tsv <- path_package("gapminder","extdata","gapminder.tsv"))


## Rectangular data 

gapminder <- read_tsv(gap_tsv)

str(gapminder, give.attr = FALSE)

gapminder <- gapminder %>%  
  mutate(country = factor(country),
         continent = factor (continent))

str(gapminder)

# Let’s create a country-level summary of maximum life expectancy

gap_life_exp <- gapminder %>% 
  group_by(country, continent) %>% 
  summarise(life_exp = max(lifeExp)) %>% 
  ungroup()

gap_life_exp


write_csv(gap_life_exp, "gap_life_exp.csv")


head(levels(gap_life_exp$country))

gap_life_exp <- gap_life_exp %>% 
  mutate(country = fct_reorder(country, life_exp)) 

head(levels(gap_life_exp$country))
  

gap_life_exp$country



### Cleaning up the files that we have created 
file.remove(list.files(pattern = "^gap_life_exp"))

```

### Chapter 10 - Factors & forcasts package 


```{r}
library(tidyverse)
library(gapminder)
str(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)

gapminder %>% 
  count(continent)

fct_count(gapminder$continent)


### Make sure to drop the factors

nlevels(gapminder$country)


h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")

h_gap <- gapminder %>% 
  filter(country %in% h_countries)

nlevels(h_gap$country)

## where we actually drop the levels of the factor 

h_gap_dropped <- h_gap %>% 
  droplevels()

nlevels(h_gap_dropped$country )


### use forcats::fct_drop() on a free- range factor 

h_gap$country %>% 
  fct_drop() %>% 
  levels()

### Filter the gapminder data down to rows where population is less than a quarter of a million.Get rid of the unused factor levels for country and continent


fct_count(gapminder$country)
fct_count(gapminder$continent)

names(gapminder)

test_gap <- gapminder %>% 
  filter( pop < 250000)

fct_count(test_gap$continent)
## should be 3 continents 

test_gap$country %>% 
fct_drop() %>% 
levels()



test_gap$continent %>% 
fct_drop() %>% 
levels()

### now trying to save it after getting it done 


test_gap_dropped <- test_gap %>% 
  droplevels()

nlevels(test_gap_dropped$continent) 
nlevels(test_gap_dropped$country)

### orderinf factor levels by frequency or by another variable 

#default

gapminder$continent %>% 
  levels()

## ordering by frequency 

gapminder$continent %>% 
  fct_infreq()  %>% 
  levels()

## backwards- least 
gapminder$continent %>% 
  fct_infreq() %>% 
  fct_rev %>% 
  levels()

## order countries by median life expectancy

fct_reorder(gapminder$country, gapminder$lifeExp) %>%
  levels() %>% 
  head()

## order accoring to minimum life exp instead of median
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% 
  levels() %>%
  head()
## backwards (highest life exp)
fct_reorder(gapminder$country, gapminder$lifeExp, .desc = TRUE) %>% 
  levels() %>%
  head()

###### Example of why we reorder factor levels: often makes plots much better! When a factor is mapped to x or y, it should almost always be reordered by the quantitative variable you are mapping to the other one

 gap_asia_2007 <- gapminder %>% 
  filter(year == 2007 , continent == "Asia")

 ggplot(gap_asia_2007,aes(x = lifeExp, y = country)) + geom_point()
 
ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) + geom_point()



#Use fct_reorder2() when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. This way the legend appears in some order as the data!

h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")

h_gap <- gapminder %>% 
  filter(country %in% h_countries) %>% 
  droplevels()

ggplot(h_gap, aes(x= year, y = lifeExp, color = country)) + geom_line()

ggplot(h_gap, aes(x= year, y = lifeExp, color = fct_reorder2(country, year, lifeExp))) + geom_line() + labs(color = "country")





### Stopped at 10.7 too tired to continue ....hmmm..hmmm

## New day Jan 22-2020 - Growing a factor 
df1 <- gapminder %>% 
  filter(country %in% c("United States", "Mexico"), year > 2000) %>% 
  droplevels()

df2 <- gapminder %>% 
  filter(country %in% c("France","Germany"), year > 2000) %>% 
  droplevels()

levels(df1$country)

levels (df2$country)
## combining using fct_c

fct_c(df1$country, df2$country)

```

 ## Chapter 11 - Character vector 
 
 Stringr package - manipulating character vectors 
 
 Especially useful for functions that split one character vector into many and vice versa: separate(), unite(), extract().
Base functions: nchar(), strsplit(), substr(), paste(), paste0().
The glue package is fantastic for string interpolation. If stringr::str_interp() doesn’t get your job done, check out the glue package.

```{r}
## Detect or filter on a target string - str_

str_detect(fruit, pattern = "fruit")
fruit


## Only keep the actual fruits that match  

my_fruit <- str_subset(fruit, pattern = "fruit")
my_fruit

## substring extraction (and replacement by position)

length (my_fruit)

str_length(my_fruit)
## first 3 letters
head(fruit) %>% 
  str_sub(1,3) 

### The regex a.b will match all countries that have an a, followed by any single character, followed by b. Yes, regexes are case sensitive, i.e. “Italy” does not match.
countries <- levels(gapminder$country)
str_subset(countries, pattern = "i.a")

## Anchors can be included to express where the expression must occur within the string. The ^ indicates the beginning of string and $ indicates the end.

str_subset(countries, pattern = "i.a$")

str_subset(my_fruit, pattern = "d")

str_subset(my_fruit, pattern = "^d")

```
 
 
 ## 13 - Dates and Times 
 
```{r}
library(tidyverse)
library(lubridate)
Sys.Date()
today()

str(Sys.Date())
class(Sys.Date())
str(today())
class(today())


Sys.time()

now()

str(Sys.time())

class(Sys.time())

str(now())

class(now())




### Column binding 

library(gapminder)

life_exp <- gapminder %>% 
  select(country, year,lifeExp)

pop  <- gapminder %>% 
  arrange(year) %>%
  select(pop)

gdp_percap <- gapminder %>% 
  arrange(pop) %>% 
  select(gdpPercap)

(gapminder_garbage <- bind_cols(life_exp, pop,gdp_percap))


summary(gapminder$lifeExp)

summary(gapminder_garbage$lifeExp)

range(gapminder$gdpPercap)

range(gapminder_garbage$gdpPercap)
```
 
 #### 18 - Writing your own functions, part 1
 
 
 Build that skateboard before you build the car or some fancy car part. A limited-but-functioning thing is very useful. It also keeps the spirits high
 
 
 The special argument ... is very useful when you want the ability to pass arbitrary arguments down to another function, but without constantly expanding the formal arguments to your function. This leaves you with a less cluttered function definition and gives you future flexibility to specify these arguments only when you need to.
 
 
 Until now, we’ve relied on informal tests of our evolving function. If you are going to use a function a lot, especially if it is part of a package, it is wise to use formal unit tests.

The testthat package (CRAN; GitHub) provides excellent facilities for this, with a distinct emphasis on automated unit testing of entire packages. However, we can take it out for a test drive even with our one measly function.

We will construct a test with test_that()
 
 
```{r}
library(gapminder)
str(gapminder)

min(gapminder$lifeExp)
max(gapminder$lifeExp)
range(gapminder$lifeExp)


### first function


max_minus_min <- function(x) max(x) - min(x)
max_minus_min(gapminder$lifeExp)
max_minus_min(1:10)
max_minus_min(runif(1000))

### stop if not 

mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
} 
mmm(gapminder)


mmm2 <- function(x){
  if(!is.numeric(x)) {
    stop('I am so sorry, but this function only works for numeric input!\n',
         'You have provided an object of class: ', class(x)[1])
  } 
  max(x)- min(x)
}
mmm2(gapminder)



quantile(gapminder$lifeExp)

quantile(gapminder$lifeExp, probs = .50)

median(gapminder$lifeExp)

quantile(gapminder$lifeExp, probs = c(.25,.75))

boxplot(gapminder$lifeExp, plot = FALSE)$stats

### Now write a code snippet that takes the difference between two quantiles.

the_probs <- c(.25,.75)
the_quantiles <- quantile(gapminder$lifeExp, probs = the_probs )
max(the_quantiles) - min(the_quantiles)


qdiff1 <- function(x,probs) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile( x = x , probs = probs)
  max(the_quantiles) - min(the_quantiles)
}


qdiff1(gapminder$lifeExp, probs = c(.25,.75))

IQR(gapminder$lifeExp)



qdiff3 <- function(x, probs = c(0, 1)) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs)
  max(the_quantiles) - min(the_quantiles)
}
qdiff3(gapminder$lifeExp)


z <- gapminder$lifeExp
z[3] <- NA
quantile(gapminder$lifeExp)

quantile(z)

quantile(z, na.rm = TRUE)

### Dealing with NA'S within our quantile function 

qdiff4 <- function(x, probs = c(0, 1)) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs, na.rm = TRUE)
  max(the_quantiles) - min(the_quantiles)
}
qdiff4(gapminder$lifeExp)

qdiff4(z)

### Dont give defaults, on this function we set na.rm = na.rm 

qdiff5 <- function( x, probs = c(0,1), na.rm = TRUE) {
  stopifnot(is.numeric(x))
  the_quantiles <- quantile(x, probs, na.rm = na.rm)
  max(the_quantiles) - min(the_quantiles)
}
  
qdiff5(gapminder$lifeExp)

qdiff5(z)

 qdiff5(z, na.rm = FALSE) 
 
 
 #### we can use ellipsis in function to give the end user freedom to add to the function
 
 qdiff6 <- function( x, probs = c(0,1), na.rm = TRUE, ...) {
   stopifnot(is.numeric(x)) 
   the_quantiles <- quantile(x, probs, na.rm = na.rm )
   max(the_quantiles) - min(the_quantiles)
 }


 
### showing difference between type 1 and type 2  
set.seed(1234)
z <- rnorm(10)
quantile(z, type = 1)
quantile(z, type = 3)
all.equal(quantile(z, type = 1), quantile(z,type = 3))

 qdiff6(z, probs = c(0.25,.075), type = 1)
 
 
  qdiff6(z, probs = c(0.25,.075), type = 3)
  
  
  
#### Let’s revert to a version of our function that does no NA handling, then test for proper NA handling. We can watch it fail.
  
library(testthat)
 
qdiff_no_NA_no <- function(x, probs = c(0, 1)) {
  the_quantiles <- quantile(x = x, probs = probs)
  max(the_quantiles) - min(the_quantiles)
}

### doesnt work and test that detects it correctly 
test_that('Na handling works' , {
  expect_that(qdiff_no_NA_no(c(1:5,NA)), equals(4))
})
##works 
qdiff_no_NA <- function(x, probs = c(0, 1), na.rm = TRUE) {
  the_quantiles <- quantile(x = x, probs = probs, na.rm = na.rm)
  max(the_quantiles) - min(the_quantiles)
}


test_that('Na handling works' , {
  expect_that(qdiff_no_NA(c(1:5,NA)), equals(4))
})
  
```

## Chapter 21 - Function writing practicum 
 
```{r}
library(gapminder)
library(ggplot2)
library(dplyr)

## getting data to practice 
j_country <- "France"
(j_dat <- gapminder %>% 
  filter(country == j_country))

## Always always always plot the data 

p <- ggplot(j_dat, aes(x = year, y = lifeExp))
p + geom_point() + geom_smooth(method = "lm", se = FALSE)


##Fit the regression  - sanity check - change the intercept 

j_fit <- lm(lifeExp ~ year , j_dat)
coef(j_fit)

## coef didnt make sense because it says at year 0 life expectancy was at -397.76 so we will move it to start that the beginning of the data 1952....I() function which “inhibits interpretation/conversion of objects”

j_fit <- lm(lifeExp ~ I(year - 1952), j_dat)
coef(j_fit)
### Coefficient of 67.79 makes much more sense 

### turn working code into a function 

le_lin_fit <- function ( dat, offset = 1952) {
  the_fit <- lm(lifeExp ~ I(year - offset), dat)
  coef(the_fit)
}

le_lin_fit(j_dat)


## its messy and we need to fix the names on it 

le_lin_fit <- function ( dat, offset = 1952) {
  the_fit <- lm(lifeExp ~ I(year - offset), dat)
  setNames(coef(the_fit), c("intercept", "slope"))
}
le_lin_fit(j_dat)



### testing the function on another country 

j_country <- "Zimbabwe"

(j_dat <- gapminder %>% 
  filter( country == j_country))

### visualizing 

p <- ggplot(j_dat, aes(x = year, y = lifeExp)) + geom_point()
p + geom_smooth(method = "lm", se = FALSE)

## coefficients 

j_fit <- lm(lifeExp ~ year , j_dat)
coefficients(j_fit)

### create a function 

le_lin_fit(j_dat)



### It’s also a good idea to clean out the workspace, rerun the minimum amount of code, and retest your function. This will help you catch another common mistake: accidentally relying on objects that were lying around in the workspace during development but that are not actually defined in your function nor passed as formal arguments.

rm(list = ls())

le_lin_fit <- function(dat, offset = 1952) {
  the_fit <- lm(lifeExp ~ I(year - offset), dat)
  setNames(coef(the_fit), c("Intercept", "Slope"))
}

le_lin_fit(gapminder %>% filter(country == "Zimbabwe"))
le_lin_fit(gapminder %>% filter(country == "Kenya"))
```

 
 
 