---
title: "Stat545"
author: "Mark Kinyanjui"
date: "1/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning new techniques Stat 545

https://stat545.com/basic-data-care.html


Laoding all the Packages needed

```{r}
install.packages("gapminder")
install.packages("tidyverse")
library(gapminder)
library(tidyverse)


```


## Exploring the structures and datasets 

Different ways to explore a data frame, tibble 

```{r}
str(gapminder)

class(gapminder)

gapminder

head(gapminder)

tail(gapminder)

as_tibble(iris)

names(gapminder)

summary(gapminder)


# Simple plot jus to visualize dataset 

plot(lifeExp ~ year, gapminder)


plot(lifeExp ~ gdpPercap, gapminder)

plot(lifeExp ~ log(gdpPercap), gapminder)

head(gapminder$lifeExp)

summary(gapminder$lifeExp)

hist(gapminder$lifeExp)

class(gapminder$continent)

summary(gapminder$continent)

levels(gapminder$continent)


nlevels(gapminder$continent)


```

# Slowly diving deeper into the learning 

```{r}
table(gapminder$continent)

barplot(table(gapminder$continent))


```

# Using Dplyr  
It is part of the tidyverse package,

Cmd+Shift+M (Mac).- shortcut for %>%

```{r}
filter(gapminder, lifeExp < 29)

filter(gapminder, country == "Rwanda" , year >1979)

filter(gapminder, country %in% c("Rwanda", "Afghanistan"))


gapminder %>% 
  select(year, lifeExp) %>% 
  head(4)

gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(year, lifeExp)


### Creating a copy of gapminder so no changes are made to the original dataset

(my_gap <- gapminder)

my_gap %>% 
  mutate(gdp = pop * gdpPercap)

```

### Replicating data in a row to fit number of levels 
Maybe it would be more meaningful to consumers of my tables and figures to stick with GDP per capita. But what if I reported GDP per capita, relative to some benchmark country. Since Canada is my adopted home, I’ll go with that.

I need to create a new variable that is gdpPercap divided by Canadian gdpPercap, taking care that I always divide two numbers that pertain to the same year.

How I achieve this:

1.Filter down to the rows for Canada.
2.Create a new temporary variable in my_gap:
3.Extract the gdpPercap variable from the Canadian data.
4.Replicate it once per country in the dataset, so it has the right length.
5.Divide raw gdpPercap by this Canadian figure.
6.Discard the temporary variable of replicated Canadian gdpPercap



summarize_at() applies the same summary function(s) to multiple variables

```{r}
my_gap <- gapminder
ctib <- my_gap %>% 
  filter(country == "Canada") 

my_gap <- my_gap %>% 
  mutate (tmp = rep(ctib$gdpPercap, nlevels(country)),
         gdpPercapRel = gdpPercap / tmp,
         tmp = NULL)
head(my_gap)

summary (my_gap$gdpPercapRel)


### Select 

my_gap %>% 
  filter(country == "Burundi", year > 1996) %>% 
  select( yr = year, lifeExp, gdpPercap) %>% 
  select(gdpPercap, everything())


#How many observations do we have per continent?
my_gap %>% 
  group_by(continent) %>% 
  summarize(n = n())

# The tally() function is a convenience function that knows to count rows. It honors groups

my_gap %>% 
  group_by(continent) %>% 
  tally()

# The count() function is an even more convenient function that does both grouping and counting

my_gap %>% 
  count(continent)


#What if we wanted to add the number of unique countries for each continent?

my_gap %>% 
  group_by(continent) %>% 
  summarize ( n = n(),
              n_countries = n_distinct(country)) 

my_gap %>% 
  group_by(continent) %>% 
  summarize(avg_lifeExp = mean(lifeExp))


# Let’s compute average and median life expectancy and GDP per capita by continent by year…but only for 1952 and 2007.
my_gap %>% 
  filter(year %in% c(1952,2007)) %>% 
  group_by(continent, year) %>% 
  summarize_at(vars(lifeExp,gdpPercap), list(~mean(.),~median(.)))


#Let’s focus just on Asia. What are the minimum and maximum life expectancies seen by year?

my_gap %>% 
  filter(continent == "Asia") %>% 
  group_by(year) %>% 
  summarize ( min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp))

```
# Computing with group-wise summaries
Let’s make a new variable that is the years of life expectancy gained (lost) relative to 1952, for each individual country. We group by country and use mutate() to make a new variable. The first() function extracts the first value from a vector. Notice that first() is operating on the vector of life expectancies within each country group.

```{r}
my_gap %>% 
  group_by(country) %>% 
  select(country, year, lifeExp) %>% 
  mutate(lifeExp_gain = lifeExp - first(lifeExp)) %>% 
  filter(year < 1963)


my_gap %>%
  filter(continent == "Asia") %>% 
  select(year, country, lifeExp) %>% 
  group_by(year) %>% 
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2 ) %>% 
  arrange(year) %>% 
  print(n = Inf)


##  which country experienced the sharpest 5-year drop in life expectancy

my_gap %>% 
  select(country, year, continent, lifeExp) %>%
  group_by(continent, country) %>%
  mutate( le_delta = lifeExp - lag(lifeExp)) %>%
  summarize( worst_le_delta = min(le_delta, na.rm = TRUE)) %>% 
  top_n(-1 , wt = worst_le_delta) %>% 
  arrange(worst_le_delta)



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
